---
title: "Data Export"
subtitle: Dataset files exported from the pilot database
---

```{css, echo=FALSE}
th {
  white-space: nowrap;
}
```

Three data files were created for the initial analysis using the Vannmiljø data.

## Data Availability

Three tab-delimited files containing all major data from the database, particle fractions, and major data plus particle fractions are available for download at the following location.

 * [pilot_vannmiljo_all.tsv.gz](https://github.com/seafood-hazards/vannmiljo-pilot/releases/download/v0.1.16/pilot_vannmiljo_all.tsv.gz)
 * [pilot_vannmiljo_particles.tsv.gz](https://github.com/seafood-hazards/vannmiljo-pilot/releases/download/v0.1.16/pilot_vannmiljo_particles.tsv.gz)
 * [pilot_vannmiljo_pivoted_particles.tsv.gz](https://github.com/seafood-hazards/vannmiljo-pilot/releases/download/v0.1.16/pilot_vannmiljo_pivoted_particles.tsv.gz)

::: {.callout-important}
The `pilot_vannmiljo_pivoted_particles.tsv.gz` file is the recommended starting point for sediment chemistry analyses.
:::

## All Major Data

The file `pilot_vannmiljo_all.tsv.gz` is a denormalized (flat) export of the relational database, designed to facilitate immediate analysis without requiring complex SQL joins.

### Column Definition

```{r}
#| echo: false
library(tibble)

tribble(
  ~`Column Name`,    ~`Source Table`,            ~Description,
  "activity_id",     "Activity",                 "Unique identifier for the activity/monitoring program",
  "activity_name",   "Activity",                 "Name of the activity or monitoring program",
  "site_code",       "Site",                     "Unique code for the sampling site",
  "site_name",       "Site",                     "Name of the sampling location",
  "lat",             "Site",                     "Latitude coordinate (decimal degrees)",
  "lon",             "Site",                     "Longitude coordinate (decimal degrees)",
  "dist_to_coast",   "Site",                     "Distance to the nearest coastline",
  "country",         "Site",                     "Country name",
  "country_code",    "Site",                     "Country code",
  "municipality",    "Site",                     "Municipality name",
  "sea_name",        "Site",                     "Name of the sea or coastal water body",
  "sample_time",     "Sample",                   "Date and time when the sample was collected",
  "upper_depth",     "Sample",                   "Upper depth of the sampled interval",
  "lower_depth",     "Sample",                   "Lower depth of the sampled interval",
  "sample_no",       "Sediment",                 "Physical sample identification number",
  "n_values",        "Sediment",                 "Number of values or replicates used to derive the measurement",
  "param_id",        "Parameter",                "Unique identifier for the chemical parameter",
  "param_name",      "Parameter",                "Name of the chemical parameter measured",
  "method",          "Sample Method",            "Method or equipment used to take the sample",
  "analysis",        "Analysis Method",          "Laboratory analysis method used",
  "value",           "Sediment",                 "Measured concentration or numerical result",
  "unit",            "Analysis Method",          "Unit of measurement for the value",
  "operator",        "Sediment",                 "Mathematical operator (e.g., <, =) indicating if value is below detection limits",
  "lod",             "LLD (Filtered)",           "Limit of Detection (LOD) value",
  "loq",             "LLD (Filtered)",           "Limit of Quantification (LOQ) value",
  "category",        "Parameter (Categorized)",  "Parameter category: efsa (EFSA elements), other (Other elements), tcs (Total sulfur and carbon), p (Particle fractions)"
)

```

## Conversion Result of Particle Fractions

The file `pilot_vannmiljo_particles.tsv.gz` contains the particle size conversion results. For a full description of the conversion methodology, see the [Sediment Particle Size Fractions](sediment-fractions.qmd) page.

### Column Definition

```{r}
#| echo: false
tribble(
  ~Column,           ~Type,       ~Description,
  "sample_id",       "character", "Unique sample identifier, linking to the broader dataset",
  "sediment_no",     "integer",   "Sediment layer number within the core (1 = topmost measured layer)",
  "clay_pct",        "numeric",   "Clay fraction (< 2 µm), as a percentage of the total sediment",
  "silt_pct",        "numeric",   "Silt fraction (2–63 µm), as a percentage",
  "sand_pct",        "numeric",   "Sand fraction (63–2000 µm), as a percentage",
  "gravel_pct",      "numeric",   "Gravel fraction (> 2000 µm), as a percentage",
  "total_pct",       "numeric",   "Sum of the four fractions; equals 100 % for all fully processed samples",
  "clay_method",     "character", "Derivation method for clay: 'direct', 'arithmetic', or 'background'",
  "silt_method",     "character", "Derivation method for silt",
  "sand_method",     "character", "Derivation method for sand",
  "gravel_method",   "character", "Derivation method for gravel",
  "any_op_adjusted", "logical",   "TRUE if any input measurement had a non-exact operator (i.e. '<', '>', or 'ND')",
  "qc_confidence",   "factor",    "Data quality confidence level: 'high', 'medium', 'low', 'very_low', or 'unreliable'"
) 

```

::: {.callout-note}
Fractions estimated using background ratios (method = `"background"`) are model-derived and carry more
uncertainty than directly measured or arithmetically derived values. The `qc_confidence`
column summarises the overall reliability of each row and should be the primary filter
applied before analysis.
:::

## Major Data with Particle Fractions

The file `pilot_vannmiljo_pivoted_particles.tsv.gz` is a combined dataset that merges the major data with the derived particle size fractions. It contains all rows from `pilot_vannmiljo_all.tsv.gz` where `category != "p"` (i.e. all chemical and other measurements, excluding the raw particle fraction rows), with the four derived grain-size fractions and their quality flag appended as additional columns. 

### Column Definition

The file contains all columns from `pilot_vannmiljo_all.tsv.gz` (see above), plus the five additional columns below.

```{r}
#| echo: false
tribble(
  ~Column,                  ~Type,     ~Description,
  "clay_pct",               "numeric", "Clay fraction (< 2 µm) derived from particle size analysis, as a percentage",
  "silt_pct",               "numeric", "Silt fraction (2–63 µm), as a percentage",
  "sand_pct",               "numeric", "Sand fraction (63–2000 µm), as a percentage",
  "gravel_pct",             "numeric", "Gravel fraction (> 2000 µm), as a percentage",
  "total_pct",              "numeric", "Sum of the four fractions; equals 100 % for all fully processed samples",
  "particle_qc_confidence", "factor",  "Data quality confidence level for the particle fractions: 'high', 'medium', 'low', 'very_low', or 'unreliable'. NA if no particle size data are available for that sample."
)

```

::: {.callout-note}
Particle fraction columns will be `NA` for samples that have no associated grain-size
measurements. The `particle_qc_confidence` column should be used to filter rows before
any analysis that depends on sediment texture.
:::
