---
title: "Sediment Particle Size Fractions"
subtitle: "From raw grain-size measurements to standardised fractions"
---

```{r}
library(tidyverse)
library(knitr)
library(kableExtra)
```

## Overview

Raw measurements for grain sizes are reported as percentages across ten size classes and aggregate measures in the Vannmiljø datasets. Because some parameters represent sub-fractions and others represent their sums (e.g. FINS spans the combined clay and silt range), the raw data cannot be used directly without first resolving these overlaps. In addition, the source data contain a number of quality issues that must be addressed before the fraction conversion can be applied.

### Particle Size Parameters in Vannmiljø

The parameters below cover the full grain-size spectrum from clay to coarse gravel in the Vannmiljø datasets. Note that `FINS` and `GSMF_63` are aggregate measures that overlap with the more detailed sub-fraction parameters above and below them.

```{r}
tribble(
  ~Parameter, ~Aggregate, ~Description,
	"GSMF2", "", "Particle fraction < 2 µm",
	"GSMF2_63", "", "Particle fraction 2 - 63 µm",
	"GSMF63_125", "", "Particle fraction 63 - 125 µm",
	"GSMF125_250", "", "Particle fraction 125 - 250 µm",
	"GSMF250_500", "", "Particle fraction 250 - 500 µm",
	"GSMF500_1000", "", "Particle fraction 500 - 1000 µm",
	"GSMF1000_2000", "", "Particle fraction 1000 - 2000 µm",
	"GSMF_2000", "", "Particle fraction > 2000 µm",
	"FINS", "✓", "Fines < 63 µm",
	"GSMF_63", "✓", "Particle fraction > 63 µm",
) |>
  kable(align = c("l", "c", "l")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

### Conversion to Four Standard Fractions

To facilitate comparison with other datasets, including the Mareano datasets, the Vannmiljø parameters are converted into four standard sediment fractions.

```{r}
tribble(
  ~Fraction, ~`Size range`, ~Description,
  "Clay",    "< 2 µm",      "Very fine particles, high surface area",
  "Silt",    "2 – 63 µm",   "Fine particles, settles slowly",
  "Sand",    "63 – 2000 µm","Medium particles comprising five sub-fractions (63–125, 125–250, 250–500, 500–1000, 1000–2000 µm)",
  "Gravel",  "> 2000 µm",   "Coarse particles > 2 mm; typically minor in open marine settings"
) |>
  kable(align = c("l", "l", "l")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

---

## Data Quality Issues

### 1. Non-standard measurement operators

Not all measurements are simple point estimates. Each value in the source data is
accompanied by an operator describing its relationship to the true value:

```{r}
tribble(
  ~Operator, ~Meaning,                                                  ~Treatment,
  "=",       "Exact measurement",                                        "Used as-is",
  "<",       "Below detection or quantification limit",                  "Replaced with half the reported value (mid-point convention)",
  ">",       "Exceeds the upper measurement range (lower bound only)",   "Used as-is (treated as a measured lower bound)"
) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

### 2. Decimal-point errors

Values well above 100 % were present in the source data, most likely caused by decimal
points being omitted or misplaced during data entry in spreadsheet software. For
example, a true value of 45.9 % was recorded as 45903.

### 3. Overlapping and redundant parameters

The source data contain both aggregate parameters and their constituent sub-fractions
for the same sample:

- **Fines < 63 µm** (`FINS`) is the sum of Clay + Silt, but some samples report
  FINS alongside separate Clay and Silt measurements, double-counting that portion.
- **Coarse > 63 µm** (`GSMF_63`) is similarly redundant when individual sand
  sub-fractions or a gravel measurement are also present.

Including both aggregate and component values inflates the apparent total well beyond
100 % and prevents straightforward fraction conversion.

---

## Processing Pipeline

The data are processed in six sequential steps. Each step feeds directly into the next.

### Step 1: Operator adjustment

Measurement operators are resolved into usable numeric values using the conventions
shown in the table above. All subsequent steps operate on these adjusted values.

### Step 1b: Decimal-point correction

For each parameter, a background median is computed from all valid values (i.e. those
already in the range 0–100 %). Values exceeding 100 % are then corrected by dividing
by successive powers of ten (10, 100, 1000, …) and selecting the candidate that is
both ≤ 100 % and closest to that parameter's background median. Values that cannot be
recovered this way are set to missing and flagged.

### Step 1c: Overlap removal and rescaling

Redundant aggregate parameters are removed from any sample where their components are
also present. Specifically:

- `FINS` is dropped if either `GSMF2` (Clay) or `GSMF2_63` (Silt) is present.
- `GSMF_63` is dropped if any sand sub-fraction or `GSMF_2000` (Gravel) is present.

After deduplication, the remaining values for each sample are summed. If the total
exceeds 100 %, all values are rescaled proportionally so that the sample sums to
exactly 100 %.

### Step 2: Wide-format restructuring

The cleaned long-format data are pivoted to wide format, giving one row per
sample × sediment layer with columns for each grain-size parameter and its operator.

### Step 3: Arithmetic derivation (four passes)

The four target fractions are derived algebraically from whichever measurements are
available, without using any background statistics. The derivation exploits the
following constraints.

 * **Clay** + **Silt** = **Fines**
 * **Sand** + **Gravel** = **Coarse**
 * **Fines** + **Coarse** = *100%*

Four passes are made in order:

| Pass | Action |
|------|--------|
| 1 | Assign directly measured values |
| 2a | Compute group totals (Fines or Coarse) from the sum of their components |
| 2b | Derive a missing component by subtracting the known one from the group total |
| 3 | Apply the 100 % constraint to derive the missing group total |
| 4 | Repeat Pass 2b using group totals that became available in Pass 3 |

This resolves the vast majority of combinations present in the dataset without
any assumptions about the sediment composition.

### Step 4:  Background ratio estimation

For samples that arithmetic derivation could not fully resolve, background ratios are
estimated from all samples where both components of a group are arithmetically known:

$$r_\text{clay/fines} = \text{median}\!\left(\frac{\text{Clay}}{\text{Clay} + \text{Silt}}\right)$$
$$\qquad r_\text{gravel/coarse} = \text{median}\!\left(\frac{\text{Gravel}}{\text{Sand} + \text{Gravel}}\right)$$

A set of overall background proportions (Clay : Silt : Sand : Gravel) is also
derived from samples where all four fractions are known, for use as a final fallback.

### Step 5: Within-group background splits

If a group total (Fines or Coarse) is known but *neither* of its components could be
derived arithmetically, the group total is split using the background ratios from Step 3.

### Step 6: General background imputation

Any fraction still unknown after Steps 2–4 is assigned a share of the remaining
percentage budget (100 % minus the sum of known fractions). The share is proportional
to that fraction's background weight. Fractions already resolved are **never modified**
at this stage.

---

## Quality Confidence Flag

Every sample receives a single `qc_confidence` rating based on accumulated penalty
points from the three pre-processing steps.

```{r}
tribble(
  ~Source,                                              ~`Penalty points`,
  "Sum of fractions 100–101 % (minor rounding)",       "+1",
  "Sum of fractions 101–110 % (moderate conflict)",    "+2",
  "Sum of fractions 110–200 % (serious conflict)",     "+3",
  "Sum of fractions > 200 % (likely multiple errors)", "+4",
  "Decimal-point correction applied (Step 0b)",        "+1",
  "Redundant aggregate removed (Step 0c)",             "+1"
) |>
  kable(align = c("l", "c")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r}
tribble(
  ~`Total penalty`, ~qc_confidence, ~Interpretation,
  "0",              "high",         "No data quality issues detected",
  "1",              "medium",       "One minor issue (e.g. rounding-only excess or one decimal correction)",
  "2",              "low",          "Two minor or one moderate issue",
  "3",              "very_low",     "Substantial conflict in the original data; values corrected by rescaling",
  "4 or more",      "unreliable",   "Severe conflict; results may not be reliable"
) |>
  kable(align = c("c", "l", "l")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  row_spec(5, bold = TRUE, color = "#8B0000")
```

Researchers are encouraged to filter on `qc_confidence` according to the requirements
of their analysis. A common starting point is to retain `"high"` and `"medium"` samples
for quantitative work, and to inspect `"low"` samples on a case-by-case basis.

---

## Derivation Method Labels

In addition to the confidence flag, each of the four output fractions carries a label
describing how it was derived:

```{r}
tribble(
  ~Method,      ~Meaning,
  "direct",     "The fraction was measured directly in the original data",
  "arithmetic", "The fraction was calculated from other measurements using mass-balance arithmetic (no assumptions about composition)",
  "background", "At least one background ratio or overall proportion was required to estimate this fraction"
) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

---

## Data Availability

A tab-delimited files containing particle fractions is available for download at the following location.

 * [pilot_vannmiljo_particles.tsv.gz](https://github.com/seafood-hazards/vannmiljo-pilot/releases/download/v0.1.5/pilot_vannmiljo_particles.tsv.gz)

### Column Definition

The final exported dataset contains one row per sample × sediment layer. The columns
are described below.

```{r}
tribble(
  ~Column,           ~Type,       ~Description,
  "sample_id",       "character", "Unique sample identifier, linking to the broader dataset",
  "sediment_no",     "integer",   "Sediment layer number within the core (1 = topmost measured layer)",
  "clay_pct",        "numeric",   "Clay fraction (< 2 µm), as a percentage of the total sediment",
  "silt_pct",        "numeric",   "Silt fraction (2–63 µm), as a percentage",
  "sand_pct",        "numeric",   "Sand fraction (63–2000 µm), as a percentage",
  "gravel_pct",      "numeric",   "Gravel fraction (> 2000 µm), as a percentage",
  "total_pct",       "numeric",   "Sum of the four fractions; equals 100 % for all fully processed samples",
  "clay_method",     "character", "Derivation method for clay: 'direct', 'arithmetic', or 'background'",
  "silt_method",     "character", "Derivation method for silt",
  "sand_method",     "character", "Derivation method for sand",
  "gravel_method",   "character", "Derivation method for gravel",
  "any_op_adjusted", "logical",   "TRUE if any input measurement had a non-exact operator (i.e. '<', '>', or 'ND')",
  "qc_confidence",   "factor",    "Data quality confidence level: 'high', 'medium', 'low', 'very_low', or 'unreliable'"
) |>
  kable(align = c("l", "l", "l")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) |>
  column_spec(1, bold = TRUE, monospace = TRUE) |>
  column_spec(2, monospace = TRUE)
```

::: {.callout-note}
Fractions estimated using background ratios (method = `"background"`) are model-derived and carry more
uncertainty than directly measured or arithmetically derived values. The `qc_confidence`
column summarises the overall reliability of each row and should be the primary filter
applied before analysis.
:::
